---
title: "FitCurve"
author: "QianqianHuang"
date: "2025-05-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(dplyr)
library(tibble)
library(magrittr)
library(minpack.lm)
library(nlme)
library(brms)
library(ggplot2)
library(lme4)
library(lmerTest)   # for calculate p-value in model with random effects
library(performance)
library(broom)      # for glance()
library(loo)        # for LOO-AIC (brms)
library(emmeans)
```

## read file

```{r, include=TRUE, echo=TRUE}
data <- read.csv("../sow_FI_weather.csv", header = T)
#data_P1 <- data %>%
#  filter(Parity == "P1") %>%
#  select(SowID, Day, FI) %>%
#  filter(!is.na(FI), !is.na(Day), !is.na(SowID)) %>%
#  mutate(SowID = factor(SowID)) 
```

## check how many data in each lactation day

```{r, include=TRUE, echo=TRUE}
print(data %>%
  group_by(Day, Parity) %>%
  summarise(
    Count = n(),
    Mean_FI = mean(FI, na.rm = TRUE),
    SD_FI = sd(FI, na.rm = TRUE)
  ), n=100)
```
Since the record from Day 23-25 lower than 10, remove them.

```{r, include=TRUE, echo=TRUE}
data_trimmed <- data %>%
  filter(Day <= 22)
```

Plot the original mean pattern.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
data_trimmed %>%
  filter(Parity == "P1")  %>%
  ggplot(aes(x = Day, y = FI)) +
  geom_jitter(alpha = 0.2, width = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(title = "Original pattern of P1") +
  theme_minimal()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
data_trimmed %>%
  filter(Parity == "P2")  %>%
  ggplot(aes(x = Day, y = FI)) +
  geom_jitter(alpha = 0.2, width = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(title = "Original pattern of P2") +
  theme_minimal()
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
data_trimmed %>%
  filter(Parity == "P3+")  %>%
  ggplot(aes(x = Day, y = FI)) +
  geom_jitter(alpha = 0.2, width = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  labs(title = "Original pattern of P3+") +
  theme_minimal()
```



## ---- Method 1: GMM -----##
**DFIi,t = DFI0+[(DFIA-DFI0)(t/k)^c]/[1+(t/k)^c]**
DFI0: predicted DFI at day 0, DFIA is asymptotic DFI


**1. Fixed-effects model**

```{r, include=TRUE, echo=TRUE}
form1 <- bf(
  FI ~ DFI0 + ((DFIA - DFI0) * (Day / k)^c) / (1 + (Day / k)^c),
  DFI0 + DFIA + k + c ~ 1,
  nl = TRUE
)
## set prior, can use get_prior(form, data_trimmed) function to check the parameter name
priors1 <- c(
  # DFI0 (>= 0)
  prior(normal(3, 2), class = "b", nlpar = "DFI0", lb = 0),
  # DFIA (>= 0)
  prior(normal(12, 3), class = "b", nlpar = "DFIA", lb = 0),
  # k (>= 0.1)
  prior(normal(6, 2), class = "b", nlpar = "k", lb = 0.1),
  # c (>= 0.5)
  prior(normal(1.5, 0.5), class = "b", nlpar = "c", lb = 0.5)
)

## run model
GMM_fixed <- brm(
  formula = form1,
  data = data_trimmed,
  family = gaussian(),
  prior = priors1,
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  seed = 123
)
```


```{r, include=TRUE, echo=TRUE}
summary(GMM_fixed)
```

**2. Model with DFI0 as a random effect**

Start value get from the coef of previous fixed-effects model.
Add lower bound to avoid bad prior sample.

```{r, include=TRUE, echo=TRUE}
form2 <- bf(
  FI ~ DFI0 + ((DFIA - DFI0) * (Day / k)^c) / (1 + (Day / k)^c),
  DFI0 ~ 1 + (1 | SowID),
  DFIA + k + c ~ 1,
  nl = TRUE
)

priors2 <- c(
   prior(normal(7.6, 0.4), class = "b", nlpar = "DFI0", lb = 0),   # DFI0 fixed
   prior(normal(21.4, 1.0), class = "b", nlpar = "DFIA", lb = 0),
   prior(normal(9.7, 0.5), class = "b", nlpar = "k", lb = 0.1),
   prior(normal(2.3, 0.3), class = "b", nlpar = "c", lb = 0.5),
   
   prior(exponential(1), class = "sd", nlpar = "DFI0", group = "SowID"),
   prior(exponential(1), class = "sigma")
)
```


```{r, include=TRUE, echo=TRUE}
GMM_dfi0 <- brm(
  formula = form2,
  data = data_trimmed,
  family = gaussian(),
  prior = priors2,
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  seed = 123
)
summary(GMM_dfi0)
```

**3. Model with DFIA as a random effect**

```{r, include=TRUE, echo=TRUE}
form3 <- bf(
  FI ~ DFI0 + ((DFIA - DFI0) * (Day / k)^c) / (1 + (Day / k)^c),
  DFIA ~ 1 + (1 | SowID),
  DFI0 + k + c ~ 1,
  nl = TRUE
)

priors3 <- c(
   prior(normal(7.6, 0.4), class = "b", nlpar = "DFI0", lb = 0),   # DFI0 fixed
   prior(normal(21.4, 1.0), class = "b", nlpar = "DFIA", lb = 0),
   prior(normal(9.7, 0.5), class = "b", nlpar = "k", lb = 0.1),
   prior(normal(2.3, 0.3), class = "b", nlpar = "c", lb = 0.5),
   
   prior(exponential(1), class = "sd", nlpar = "DFIA", group = "SowID"),
   prior(exponential(1), class = "sigma")
)
```


```{r, include=TRUE, echo=TRUE}
GMM_dfiA <- brm(
  formula = form3,
  data = data_trimmed,
  family = gaussian(),
  prior = priors3,
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  seed = 123
)
summary(GMM_dfiA)
```



**4. Model with DFI0, DFIA as a random effect**

```{r, include=TRUE, echo=TRUE}
form4 <- bf(
  FI ~ (DFI0 + dfi0) + (((DFIA + dfiA) - (DFI0 + dfi0)) * (Day / k)^c) / (1 + (Day / k)^c),
  dfi0 + dfiA ~ 0 + (1 | p | SowID),    ## (1 | p | SowID) can estimate covariance
  DFI0 + DFIA + k + c ~ 1,
  nl = TRUE
)

priors4 <- c(
   prior(normal(7.6, 0.4), class = "b", nlpar = "DFI0", lb = 0),   # DFI0 fixed
   prior(normal(21.4, 1.0), class = "b", nlpar = "DFIA", lb = 0),
   prior(normal(9.7, 0.5), class = "b", nlpar = "k", lb = 0.1),
   prior(normal(2.3, 0.3), class = "b", nlpar = "c", lb = 0.5),
   
   prior(exponential(1), class = "sd", nlpar = "dfi0", group = "SowID"),
   prior(exponential(1), class = "sd", nlpar = "dfiA", group = "SowID"),
   prior(exponential(1), class = "sigma")
)
```


```{r, include=TRUE, echo=TRUE}
GMM_dfi0A <- brm(
  formula = form4,
  data = data_trimmed,
  family = gaussian(),
  prior = priors4,
  chains = 4, cores = 2,
  control = list(adapt_delta = 0.95),
  seed = 123
)
summary(GMM_dfi0A)
VarCorr(GMM_dfi0A, summary = TRUE)
```


**5. Model with DFI0, DFIA, k=b*DFI0 as random effects**

```{r, include=TRUE, echo=TRUE}
form5 <- bf(
  FI ~ (DFI0 + dfi0) + (((DFIA+ dfiA) - (DFI0+ dfi0)) * (Day / (k + b*dfi0))^c) / (1 + (Day / (k + b*dfi0))^c),
  DFI0 + DFIA + k + c + b ~ 1,
  dfi0 + dfiA ~ 0 + (1 | p | SowID),
  nl = TRUE
)

priors5 <- c(
   prior(normal(7.6, 0.4), class = "b", nlpar = "DFI0", lb = 0),   
   prior(normal(21.4, 1.0), class = "b", nlpar = "DFIA", lb = 0),
   prior(normal(9.7, 0.5), class = "b", nlpar = "k", lb = 0.1),
   prior(normal(2.3, 0.3), class = "b", nlpar = "c", lb = 0.5),
   prior(normal(0.3, 0.1), class = "b", nlpar = "b", lb = 0),
   
   prior(exponential(1), class = "sd", nlpar = "dfi0", group = "SowID"),
   prior(exponential(1), class = "sd", nlpar = "dfiA", group = "SowID"),
   prior(exponential(1), class = "sigma")
)
```

```{r, include=TRUE, echo=TRUE}
GMM_dfi0Ak <- brm(
  formula = form5,
  data = data_trimmed,
  family = gaussian(),
  prior = priors5,
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  seed = 123
)
summary(GMM_dfi0Ak)
```

**6. Model with DFI0, DFIA, k=b*DFA0 as random effects**

```{r, include=TRUE, echo=TRUE}
form6 <- bf(
  FI ~ (DFI0 + dfi0) + (((DFIA + dfia) - (DFI0 + dfi0)) * (Day / (k + b*dfia))^c) / (1 + (Day / (k + b*dfia))^c),
  DFI0 + DFIA + k + c + b ~ 1,
  dfi0 + dfiA ~ 0 + (1 | p | SowID),
  nl = TRUE
)

priors6 <- c(
   prior(normal(7.6, 0.4), class = "b", nlpar = "DFI0", lb = 0),   
   prior(normal(21.4, 1.0), class = "b", nlpar = "DFIA", lb = 0),
   prior(normal(9.7, 0.5), class = "b", nlpar = "k", lb = 0.1),
   prior(normal(2.3, 0.3), class = "b", nlpar = "c", lb = 0.5),
   prior(normal(0.3, 0.1), class = "b", nlpar = "b", lb = 0),
   
   prior(exponential(1), class = "sd", nlpar = "dfia", group = "SowID"),
   prior(exponential(1), class = "sd", nlpar = "dfi0", group = "SowID"),
   prior(exponential(1), class = "sigma")
)
```

```{r, include=TRUE, echo=TRUE}
GMM_dfi0Ak2 <- brm(
  formula = form6,
  data = data_trimmed,
  family = gaussian(),
  prior = priors6,
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  seed = 123
)
summary(GMM_dfi0Ak2)
```


**7. Model with DFI0, DFIA, c=b*DFI0 as random effects**

```{r, include=TRUE, echo=TRUE}
form7 <- bf(
  FI ~ (DFI0 + dfi0) + (((DFIA + dfiA) - (DFI0+ dfi0)) * (Day / k)^(c + b*dfi0)) / (1 + (Day / k)^(c + b*dfi0)),
  DFI0 + DFIA + k + c + b ~ 1,
  dfi0 + dfiA ~ 0 + (1 | p | SowID),
  nl = TRUE
)

priors7 <- c(
   prior(normal(7.6, 0.4), class = "b", nlpar = "DFI0", lb = 0),   
   prior(normal(21.4, 1.0), class = "b", nlpar = "DFIA", lb = 0),
   prior(normal(9.7, 0.5), class = "b", nlpar = "k", lb = 0.1),
   prior(normal(2.3, 0.3), class = "b", nlpar = "c", lb = 0.5),
   prior(normal(0.3, 0.1), class = "b", nlpar = "b"),
   
   prior(exponential(1), class = "sd", nlpar = "dfi0", group = "SowID"),
   prior(exponential(1), class = "sd", nlpar = "dfiA", group = "SowID"),
   prior(exponential(1), class = "sigma")
)
```

```{r, include=TRUE, echo=TRUE}
GMM_dfi0Ac <- brm(
  formula = form7,
  data = data_trimmed,
  family = gaussian(),
  prior = priors7,
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  seed = 123
)
summary(GMM_dfi0Ac)
```

**8. Model with DFI0, DFIA, c=b*DFA0 as random effects**

```{r, include=TRUE, echo=TRUE}
form8 <- bf(
  FI ~ (DFI0 + dfi0) + (((DFIA + dfia) - (DFI0 + dfi0)) * (Day / k)^(c + b*dfia)) / (1 + (Day / k)^(c + b*dfia)),
  DFI0 + DFIA + k + c + b ~ 1,
  dfi0 + dfia ~ 0 + (1 | p | SowID),
  nl = TRUE
)

priors8 <- c(
   prior(normal(7.6, 0.4), class = "b", nlpar = "DFI0", lb = 0),   
   prior(normal(21.4, 1.0), class = "b", nlpar = "DFIA", lb = 0),
   prior(normal(9.7, 0.5), class = "b", nlpar = "k", lb = 0.1),
   prior(normal(2.3, 0.3), class = "b", nlpar = "c", lb = 0.5),
   prior(normal(0.3, 0.1), class = "b", nlpar = "b"),
   
   prior(exponential(1), class = "sd", nlpar = "dfia", group = "SowID"),
   prior(exponential(1), class = "sd", nlpar = "dfi0", group = "SowID"),
   prior(exponential(1), class = "sigma")
)
```

```{r, include=TRUE, echo=TRUE}
GMM_dfi0Ac2 <- brm(
  formula = form8,
  data = data_trimmed,
  family = gaussian(),
  prior = priors8,
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  seed = 123
)
summary(GMM_dfi0Ac2)
```


### compare the fit curve VS real data

```{r, include=TRUE, echo=TRUE}
pig_id <- 183553
pig_data <- data_P1_trimmed %>% filter(SowID == pig_id)
newdata <- data.frame(
  Day = seq(min(pig_data$Day), max(pig_data$Day), length.out = 100),
  SowID = factor(pig_id, levels = levels(data_P1_trimmed$SowID))
)

pred <- fitted(fit2, newdata = newdata, re_formula = NULL, summary = TRUE)
newdata$FI_pred <- pred[, "Estimate"]
newdata$Lower <- pred[, "Q2.5"]
newdata$Upper <- pred[, "Q97.5"]

ggplot() +
  geom_line(data = pig_data, aes(x = Day, y = FI), color = "black", size = 1) +
  geom_line(data = newdata, aes(x = Day, y = FI_pred), color = "blue", size = 1) +
  # CI
  geom_ribbon(data = newdata, aes(x = Day, ymin = Lower, ymax = Upper), alpha = 0.2, fill = "blue") +
  labs(title = paste("Model Prediction vs Actual for SowID =", pig_id),
       x = "Day", y = "FI") +
  theme_minimal()
```



## ---- Method 2: polynomial -----##
**DFIi,t = t + t^2 +t^3 +t^4**

Center Day by minus mean value to avoid multicollinearity issue.

```{r, include=TRUE, echo=TRUE}
#data_P1_trimmed$Day_c <- scale(data_P1_trimmed$Day, center = TRUE, scale = FALSE)
poly_terms <- poly(data_trimmed$Day, degree = 4, raw = FALSE)
data_trimmed$t1 <- poly_terms[, 1]
data_trimmed$t2 <- poly_terms[, 2]
data_trimmed$t3 <- poly_terms[, 3]
data_trimmed$t4 <- poly_terms[, 4]
```


**1. fixed-effects model**
```{r, include=TRUE, echo=TRUE}
poly_fixed <- lm(FI ~ t1 + t2 + t3 + t4, data = data_trimmed)
summary(poly_fixed)
```

Since *t4* is not significant (P-value=0.892), remove t4.

```{r, include=TRUE, echo=TRUE}
poly_terms <- poly(data_trimmed$Day, degree = 3, raw = FALSE)
data_trimmed$t1 <- poly_terms[, 1]
data_trimmed$t2 <- poly_terms[, 2]
data_trimmed$t3 <- poly_terms[, 3]
poly_fixed <- lm(FI ~ t1 + t2 + t3, data = data_trimmed)
summary(poly_fixed)
```


**2. random model with intercept as random**

```{r, include=TRUE, echo=TRUE}
poly_r1 <- lmer(FI ~ t1 + t2 + t3 + (1 | SowID), data = data_trimmed)
summary(poly_r1)
```

**3. random model with intercept and t1 as random**

```{r, include=TRUE, echo=TRUE}
poly_r2 <- lmer(FI ~ t1 + t2 + t3 + (1 + t1 | SowID), data = data_trimmed)

## refit by using another optimizer
poly_r2_refit <- update(poly_r2, control = lmerControl(optimizer = "bobyqa"))
summary(poly_r2_refit)
```

**4. random model with intercept and t1,t2 as random**

```{r, include=TRUE, echo=TRUE}
poly_r3 <- lmer(FI ~ t1 + t2 + t3 + (1 + t1 + t2 | SowID), data = data_trimmed)
summary(poly_r3)
```


## compare the model

```{r, include=FALSE, echo=FALSE}
GMM_models <- list(GMM_fixed, GMM_dfi0, GMM_dfiA, GMM_dfi0A, GMM_dfi0Ak, GMM_dfi0Ak2, GMM_dfi0Ac, GMM_dfi0Ac2)
##LOOIC for Bayesian model
loo_list <- lapply(GMM_models, function(m) loo(m, moment_match = TRUE,mc.cores=4)) 
r2_list <- lapply(GMM_models, bayes_R2)
waic_list <- lapply(GMM_models, waic)
rsd_list <- lapply(GMM_models, function(mod) {
  posterior_summary(mod, variable = "sigma")["sigma", "Estimate"]  # RSD
})
```


```{r, include=TRUE, echo=TRUE}
model_summary <- data.frame(
  Model = c("GMM_fixed", "GMM_dfi0", "GMM_dfiA", "GMM_dfi0A", "GMM_dfi0Ak", "GMM_dfi0Ak2", "GMM_dfi0Ac", "GMM_dfi0Ac2"),
  WAIC =  sapply(waic_list, function(w) w$estimates["waic", "Estimate"]),
  R2 = sapply(r2_list, function(x) x[1, "Estimate"]),
  RSD = unlist(rsd_list)
)

print(model_summary)
```

Extract the SE of Var(e).

```{r, include=TRUE, echo=TRUE}
extract_var_e <- function(mod) {
  draws <- as_draws_df(mod)
  
  if (!"sigma" %in% colnames(draws)) {
    warning("No 'sigma' parameter found in model.")
    return(data.frame(Var_e = NA, SE = NA))
  }

  sigma_draws <- draws$sigma
  var_e_draws <- sigma_draws^2
  
  data.frame(
    Var_e = mean(var_e_draws),
    SE = sd(var_e_draws)
  )
}
var_e_results <- do.call(rbind, lapply(GMM_models, extract_var_e))
var_e_results$Model <- c("GMM_fixed", "GMM_dfi0", "GMM_dfiA", "GMM_dfi0A", "GMM_dfi0Ak", "GMM_dfi0Ak2", "GMM_dfi0Ac", "GMM_dfi0Ac2")
print(var_e_results)
```





```{r, include=TRUE, echo=TRUE}
poly_models <- list(poly_fixed, poly_r1, poly_r2_refit, poly_r3)
r2_list <- lapply(poly_models, r2)

poly_model_summary <- data.frame(
  Model = c("poly_fixed", paste0("poly_r", 1:3)),
  AIC = sapply(poly_models, AIC),
  R2 = sapply(seq_along(r2_list), function(i) {
    r2 <- r2_list[[i]]
    if (!is.null(r2$R2)) {
      return(r2$R2)  # 固定效应模型
    } else if (!is.null(r2$R2_conditional)) {
      return(r2$R2_conditional)  # 随机效应模型
    } else {
      return(NA)
    }
  }),
  RSD = sapply(poly_models, function(m) {summary(m)$sigma})
)
poly_model_summary
```

Extract the SE of Var(e).

```{r, include=TRUE, echo=TRUE}
extract_var_e_lm <- function(model_list, model_names = NULL) {
  results <- lapply(model_list, function(m) {
    sigma <- summary(m)$sigma
    n <- nobs(m)
    p <- length(coef(m))
    df_residual <- n - p

    var_e <- sigma^2
    se_var_e <- var_e * sqrt(2 / df_residual)

    data.frame(Var_e = var_e, SE = se_var_e)
  })

  results_df <- do.call(rbind, results)
  if (!is.null(model_names)) {
    results_df$Model <- model_names
  } else {
    results_df$Model <- paste0("Model_", seq_along(model_list))
  }

  return(results_df)
}
residual_var_results <- extract_var_e_lm(poly_models, 
                                         c("poly_fixed", "poly_r1", "poly_r2_refit", "poly_r3"))
print(residual_var_results)
```




## Calculate the mean effect group by Parity

*1. Based on best poly model: poly_r3 <- lmer(FI ~ t1 + t2 + t3 + (1 + t1 + t2 | SowID), data = data_trimmed)

```{r, include=TRUE, echo=TRUE}
fix_eff <- fixef(poly_r3)
ran_eff <- ranef(poly_r3)$SowID %>%
  rownames_to_column(var = "SowID")

sow_parity <- data_trimmed %>%
  select(SowID, Parity) %>%
  mutate(SowID = as.character(SowID)) %>%
  distinct()

individual_effects <- ran_eff %>%
  mutate(
    Intercept_total = `(Intercept)` + fix_eff["(Intercept)"],
    t1_total = t1 + fix_eff["t1"],
    t2_total = t2 + fix_eff["t2"]
  ) %>%
  select(SowID, Intercept_total, t1_total, t2_total) %>%
  left_join(sow_parity, by = "SowID")
```

```{r, include=TRUE, echo=TRUE}
group_means <- individual_effects %>%
  group_by(Parity) %>%
  summarise(
    mean_Intercept = mean(Intercept_total),
    mean_t1 = mean(t1_total),
    mean_t2 = mean(t2_total),
    .groups = "drop"
  )
print(group_means)
```

*2. Based on best GMM model: GMM_dfi0Ac2

```{r, include=TRUE, echo=TRUE}
fixef_df <- fixef(GMM_dfi0Ac2)
DFI0_fixed <- fixef_df["DFI0_Intercept", "Estimate"]

re_summary <- as.data.frame(ranef(GMM_dfi0Ac2, summary = TRUE)$SowID[, "Estimate", ])
re_summary$SowID <- rownames(re_summary)

individual_effects2 <- re_summary %>%
  mutate(
    DFI0_total = DFI0_Intercept + DFI0_fixed,
    dfia_total = fixef_df["DFIA_Intercept", "Estimate"] + dfia_Intercept,
    C_total = fixef_df["c_Intercept", "Estimate"] + fixef_df["b_Intercept", "Estimate"] * dfia_Intercept
  ) %>%
  left_join(sow_parity, by = "SowID")
```

```{r, include=TRUE, echo=TRUE}
group_summary2 <- individual_effects2 %>%
  group_by(Parity) %>%
  summarise(
    mean_DFI0 = mean(DFI0_total, na.rm = TRUE),
    mean_dfia = mean(dfia_total, na.rm = TRUE),
    mean_C = mean(C_total, na.rm = TRUE),
    n = n()
  )
print(group_summary2)
```



```{r, include=TRUE, echo=TRUE}

```





